Drug sensitivity prediction with the normal inverse Gaussian prior informed by external data

### 2019-11-25 docs/manuscript1.Rnw
commit before analysis_gdsc.R

	\section{Simulations}
	\subsection{Setup}
	This section investigates the empirical 
	Bayes estimation properties of the model in a simulated setting; its main aim 
	is to assess hyperparameter recovery. It is a data based simulation,
	wherein the responses are simulated from a synthetic model, but the
	features are taken from the real GDSC data introducted in Section 
	\ref{sec:gdsc}. 
	The real GDSC features contain strong collinearities. Such strong
	collinearities in the design matrix impede correct parameter estimation. 
	We therefore replace the ambition of correctly estimating the $\beta_{jd}$
	with the more modest aim of approximately correct estimation of the 
	hyperparameters.

@@

	To compute performance measures, we randomly split the data into training and
	test sets, containing 
	241 cell lines each, estimate the models on the training data and compute 
	performance measures on the test data. We repeated this procedure 50 times
	and average performance measures over the splits.
	The estimated 
	models are: four models like in (\ref{eq:prior}) but with different external 
	variables, named NIG1, NIG2, NIG3, and NIG4 in the following. NIG1 does 
	not use the external covariates and estimates
	one overall $\phi$ and one overall $\chi$, NIG2 uses the external 
	covariates to estimate both $\bm{\alpha}_{\text{feat}}$ and
	$\bm{\alpha}_{\text{drug}}$, while NIG3 and NIG4 estimate only 
	$\bm{\alpha}_{\text{feat}}$ or $\bm{\alpha}_{\text{drug}}$, respectively,
	keeping either $\forall d: \tau_d^2=1$ or $\forall j,d: \gamma_{jd}^2=1$
	fixed. In addition, we estimate regular lasso and ridge models. The null model
	that fits the means of the $\y_d$ is included as a reference.

	% The bSEM model is similar to 
	% the NIG model in that it 
	% draws the $\beta_{jd}$ from a conditionally normal distribution, but 
	% fixes $\tau_d^2=1$ and draws 
	% $\gamma_{jd}^2 \sim \Gamma^{-1} (\phi_{jd}, \lambda_{jd})$, instead of 
	% from the inverse Gaussian distribution. Additionally, 
	% the bSEM model can 
	% only incorporate one binary external covariate that divides the features into 
	% two groups, i.e., $\bm{\alpha} = \begin{bmatrix} \alpha_0 & \alpha_1 
	% \end{bmatrix} \tr$ and $\bm{c}_{jd} = \begin{bmatrix} 1 & 0 \end{bmatrix} 
	% \tr$ if feature $jd$ belongs to group one and $\bm{c}_{jd} = \begin{bmatrix} 
	% 1 & 1 \end{bmatrix} \tr$ if feature $jd$ belongs to group two. Both 
	% $\bm{\alpha}_{\text{feat}}$ and $\lambda_{\text{feat}}$ are
	% estimated in a similar fashion as in Section \ref{sec:estimation}.

@@

	The model is put into practice on the GDSC data. The results in Section
	\ref{sec:gdsc} show that the inclusion of the external covariates is 
	beneficial for drug sensitivity prediction. Although the prediction MSE is
	not lower if the external covariates are included, 
	the resulting hyperparameter estimates are still informative
	and may guide future research efforts.

@@

	Although the empirical Bayes estimates of the external covariate effects 
	$\bm{\alpha}$ are informative, they are small compared to 
	bSEM's estimates. Consequently, the posteriors of the models with and without
	the external covariates (NIG2 and NIG1 in Section \ref{sec:gdsc}) do not 
	differ by much. This results similar prediction MSEs with
	the inclusion of the external covariates. This is a topic that deserves 
	further attention. One proposed solution is to include an extra 
	prior variance component that promotes a more heavy-tailed $\beta_{jd}$ prior
	distribution. An example is a variance component with a half-cauchy prior 
	distribution, i.e., the horseshoe \cite[]{carvalho_handling_2009}. 
	This solution is similar to the pInc model from 
	\cite{kpogbezan_incorporating_2019}. 
	Like the bSEM model, the pInc model gives significantly larger external
	covariate effects, 
	so we conjecture that the inclusion of a similar heavy-tailed prior variance 
	component will have a similar effect in the NIG model.

>>>>

	\section{Simulations}\ref{sec:simulations}
	\subsection{Setup}
	This section investigates the empirical 
	Bayes estimation properties of the model in a simulated setting; its main aim 
	is to assess hyperparameter recovery. It is a data based simulation,
	wherein the responses are simulated from a synthetic model, but the
	features are taken from the real GDSC data introducted in Section 
	\ref{sec:gdsc}. 
	The real GDSC features contain strong collinearities. Such strong
	collinearities in the design matrix impede correct parameter estimation. 
	We therefore replace the ambition of correctly estimating the $\beta_{jd}$
	with the more modest aim of approximately correct estimation of the 
	hyperparameters.

@@

	To compute performance measures, we randomly split the data into training and
	test sets, containing 
	241 cell lines each, estimate the models on the training data and compute 
	performance measures on the test data. We repeated this procedure 50 times
	and average performance measures over the splits.
	The estimated 
	models are: four models like in (\ref{eq:prior}) but with different external 
	variables, named NIG1, NIG2, NIG3, and NIG4 in the following. NIG1 does 
	not use the external covariates and estimates
	one overall $\phi$ and one overall $\chi$, NIG2 uses the external 
	covariates to estimate both $\bm{\alpha}_{\text{feat}}$ and
	$\bm{\alpha}_{\text{drug}}$, while NIG3 and NIG4 estimate only 
	$\bm{\alpha}_{\text{feat}}$ or $\bm{\alpha}_{\text{drug}}$, respectively,
	keeping either $\forall d: \tau_d^2=1$ or $\forall j,d: \gamma_{jd}^2=1$
	fixed.

	In addition, we estimate regular lasso and ridge models, and a
	bSEM model \cite[]{leday_gene_2017,kpogbezan_empirical_2017}. The bSEM model 
	is similar to the NIG model in that it
	draws the $\beta_{jd}$ from a conditionally normal distribution, but
	fixes $\tau_d^2=1$ and draws
	$\gamma_{jd}^2 \sim \Gamma^{-1} (\phi_{jd}, \lambda_{jd})$, instead of
	from the inverse Gaussian distribution. Additionally,
	the bSEM model can
	only incorporate one binary external covariate that divides the features into
	two groups, i.e., $\bm{\alpha} = \begin{bmatrix} \alpha_0 & \alpha_1
	\end{bmatrix} \tr$ and $\bm{c}_{jd} = \begin{bmatrix} 1 & 0 \end{bmatrix}
	\tr$ if feature $jd$ belongs to group one and $\bm{c}_{jd} = \begin{bmatrix}
	1 & 1 \end{bmatrix} \tr$ if feature $jd$ belongs to group two. Both
	$\bm{\alpha}_{\text{feat}}$ and $\lambda_{\text{feat}}$ are
	estimated in a similar fashion as in Section \ref{sec:estimation}.
	The null model that fits the means of the $\y_d$ is included as a reference.

@@

	Simulations in Section \ref{sec:simulations} show that the method is 
	reasonably able to recover hyperparameters. In addition, the simulations
	show that inclusion of informative external covariates is indeed beneficial to
	predictive performance. The model is put into practice on the GDSC data. 
	The results in Section
	\ref{sec:gdsc} show that the inclusion of the external covariates is 
	beneficial for drug sensitivity prediction. Although the prediction MSE is
	not lower if the external covariates are included, 
	the resulting hyperparameter estimates are still informative
	and may guide future research efforts.

@@

	Although the empirical Bayes estimates of the external covariate effects 
	$\bm{\alpha}$ are informative, they are small compared to 
	bSEM's estimates. Consequently, the posteriors of the models with and without
	the external covariates (NIG2 and NIG1 in Section \ref{sec:gdsc}) do not 
	differ by much. This results similar prediction MSEs with
	the inclusion of the external covariates. This is a topic that deserves 
	further attention. One proposed solution is to include a
	prior variance component that promotes a more heavy-tailed $\beta_{jd}$ prior
	distribution. An example is a variance component with a half-cauchy prior 
	distribution, i.e., the horseshoe \cite[]{carvalho_handling_2009}. 
	This solution is similar to the pInc model from 
	\cite{kpogbezan_incorporating_2019}. 
	Like the bSEM model, the pInc model gives significantly larger external
	covariate effects, 
	so we conjecture that the inclusion of a similar heavy-tailed prior variance 
	component will have a similar effect in the NIG model.
